---
title: "Categorial Variables and Factors Assignment"
author: "Tyler Kephart"
date: "2024-08-18"
output: pdf_document
geometry: "margin=1.5cm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loaded libraries

```{r libraries}
library(tidyverse)
```

### 1. From the "forcats" library load gss_cat data. Explore the distribution of rincome (reported income). What makes the default bar chart hard to understand? How could you improve the plot?

```{r question-1-response}
# default bar chart of rincome
gss_cat %>% 
    ggplot(aes(x = rincome)) +
    geom_bar() +
    labs(x = "Reported Income",
         y = "Count")

# The default bar chart can be hard to understand due to:
# - overlapping labels
# - order of categories

# improved plot
gss_cat %>% 
    # reorder categories by frequency
    ggplot(aes(x = fct_infreq(rincome))) +
    geom_bar() +
    labs(x = "Reported Income",
         y = "Count") +
    # angle x variable names for readability
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 2. What is the most common religion? What is the most common partyid?

```{r question-2-response}
# most common religion
gss_cat %>%
    count(relig) %>%
    arrange(desc(n)) %>%
    slice(1)
# most common partyid
gss_cat %>%
    count(partyid) %>%
    arrange(desc(n)) %>%
    slice(1)
```

### 3. Which relig does denom (denomination) apply to? How can you find out with a table? How can you find out with a visualisation?

```{r question-3-response}
# summary table of denominations per religion
gss_cat %>%
    group_by(relig) %>%
    # counts distinct denominations for each religion
    summarise(count_denom = n_distinct(denom)) %>%
    arrange(desc(count_denom))
# plot of denominations per religion
gss_cat %>%
    group_by(relig) %>%
    # counts distinct denominations for each religion
    summarise(count_denom = n_distinct(denom)) %>%
    ggplot(aes(
            # orders religions by number of denominations
            x = reorder(relig, count_denom), 
            y = count_denom)) +
        # bar chart with height of bars being the number of denominations
        geom_col() +
        # flip the axes to make the plot easier to read
        coord_flip() +
        labs(x = "Religion",
            y = "Denominations")
```

### 4. How have the proportions of people identifying as Democrat, Republican, and Independent changed over time? Plot a suitable chart.

```{r question-4-response}
# categorize partyid into Independent, Democrat, Republican, and Other
gss_cat <- gss_cat %>%
    mutate(parties = case_when(
        partyid %in% c(
            "Ind,near rep","Independent","Ind,near dem") ~ "Independent",
        partyid %in% c(
            "Not str democrat","Strong democrat") ~ "Democrat",
        partyid %in% c(
            "Not str republican","Strong republican") ~ "Republican",
        # catch-all for any unspecified categories
        TRUE ~ "Other"),
    .after = partyid)
# summarize the proportions of parties by year
party_proportions <- gss_cat %>%
    group_by(year, parties) %>%
    summarise(count = n()) %>%
    group_by(year) %>%
    mutate(total = sum(count),
            proportion = count / total) %>%
    ungroup()
# plot the proportions of parties over time
party_proportions %>% 
    # only looking at democrat, republican, and independent, filter out others
    filter(parties != "Other") %>%
    ggplot(aes(x = year, y = proportion, color = parties)) +
        geom_line() +
        geom_point() +
        labs(x = "Year", y = "Proportion", color = "Party") +
        # make colors to match party colors
        scale_color_manual(values = c(
                            "Democrat" = "blue", 
                            "Republican" = "red", 
                            "Independent" = "black"))
```

### 5. Collapse "rincome" into smaller set of categories?

```{r question-5-response}
# define income ranges and collapse rincome into broader categories
gss_cat %>%
  mutate(income_cat = case_when(
    rincome %in% c("Lt $1000") ~ "Low Income",
    rincome %in% c("$1000 to 2999") ~ "Lower Middle Income",
    rincome %in% c("$3000 to 3999", "$4000 to 4999") ~ "Middle Income",
    rincome %in% c("$5000 to 5999", "$6000 to 6999") ~ "Upper Middle Income",
    rincome %in% c("$7000 to 7999", "$8000 to 9999") ~ "High Income",
    rincome %in% c("$10000 - 14999", "$15000 - 19999") ~ "Very High Income",
    rincome %in% c("$20000 - 24999", "$25000 or more") ~ "Top Income",
    TRUE ~ "Unknown"),
    .after = rincome)
```