---
title: "Tidying Data Assignment"
author: "Tyler Kephart"
date: "2024-08-14"
output: pdf_document
geometry: "margin=1.5cm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the tidyverse library
```{r}
library(tidyverse)
```

##### Question 1. 

The built in billboard dataset is not tidy. Describe why it is not tidy and then tidy the dataset.

```{r question-1-response}

# First gather all the week entries into a row for each week for each song
# (where there is an entry)
bb <- billboard %>% 
    pivot_longer(
        cols = starts_with("wk"), 
        names_to = "week", 
        values_to = "rank", 
        names_prefix = "wk"
        )
bb

# Then, convert the week variable to a number and...
bb$week <- parse_integer(bb$week)
# figure out the date corresponding to each week on the chart
bb <- bb %>% 
    # dropped NAs because it meant the song wasn't on the billboard
    drop_na(rank) %>% 
    mutate(date = date.entered + (7 * (week - 1)))
bb

# Sort the data by artist, track and week. 
# Here are what your first entries should be (formatting can be different):
#>  A tibble: 5,307 x 5
#   artist  track                   date.entered  week  rank   date
 #    <chr>   <chr>                   <date>       <int> <dbl>   <date>
 #  1 2 Pac   Baby Don't Cry (Keep... 2000-02-26       1    87 2000-02-26
 #  2 2 Pac   Baby Don't Cry (Keep... 2000-02-26       2    82 2000-03-04
 #  3 2 Pac   Baby Don't Cry (Keep... 2000-02-26       3    72 2000-03-11
 #  4 2 Pac   Baby Don't Cry (Keep... 2000-02-26       4    77 2000-03-18
 #  5 2 Pac   Baby Don't Cry (Keep... 2000-02-26       5    87 2000-03-25
 #  6 2 Pac   Baby Don't Cry (Keep... 2000-02-26       6    94 2000-04-01
 #  7 2 Pac   Baby Don't Cry (Keep... 2000-02-26       7    99 2000-04-08
 #  8 2Ge+her The Hardest Part Of ... 2000-09-02       1    91 2000-09-02
 #  9 2Ge+her The Hardest Part Of ... 2000-09-02       2    87 2000-09-09
 # 10 2Ge+her The Hardest Part Of ... 2000-09-02       3    92 2000-09-16
 # … with 5,297 more rows

# arranged to make sure that all the data was sorted properly
bb <- bb %>% arrange(artist, track, week)
bb
```

##### Question 2. 

Tidy the "fish_encounters" dataset of fish spotting by monitoring stations. Make the NA into 0 using the option "values_fill = list(seen = 0)"


```{r question-2-response}
fish_encounters %>% 
    complete(fish, station, fill = list(seen = 0))
# you can verify by viewing more of the data like adding "%>% print(n = Inf)"
```

##### Question 3. 

Import the flowers1 dataset. Tidy and pivot the data. Hint: use "read_csv2()" to read in the dataset 

```{r question-3-response}
# read in the data from semi-colon separated values file
flowers1 <- read_csv2("flowers1.csv")
# pivot so that the flowers value and intensity value are in each observation
flowers1 %>% pivot_wider(names_from = Variable, values_from = Value)
```

##### Question 4.

Import the flowers2 dataset. Tidy the dataset by turning the one column into 3 separate columns

```{r question-4-response}
# read in the data from semi-colon separated values file
# this will separate the time column
flowers2 <- read_csv2("flowers2.csv")
# separate the Flowers/Intensity into 2 cols and add replication col
flowers2 %>% 
    separate(
        col = `Flowers/Intensity`, 
        into = c("Flowers", "Intensity"), 
        sep = "/", 
        remove = TRUE, 
        convert = TRUE
        ) %>% 
    mutate(replication = as.numeric(rownames(.)) - (12 * (Time - 1)))
```

##### Question 5.

In the following dataset, turn the implicit missing values to explicit 

```{r question-5-response}
output <- tibble(
        treatment = c("a", "b", "a", "c", "b"),
        gender = factor(c("M", "F", "F", "M", "M"), 
        levels = c("M", "F", "O")),
        return = c(1.5, 0.75,  0.5,  1.8,  NA)
    )
output
# used complete to add the missing observations
output %>% complete(treatment, gender)
```


##### Question 6.

Import the weather dataset as weather. Use "pivot_longer()" to put the days all in one column, then use "pivot_wider" to separate tmax and tmin into separate columns. Print the summary of the final resulting dataset 

```{r question-6-response}
# read in the csv data
weather <- read_csv("weather.csv")
# pivot the days into one col
weather <- weather %>% 
    pivot_longer(
        cols = starts_with("d"), 
        names_to = "day", 
        values_to = "value", 
        names_prefix = "d", 
        names_transform = list(day = as.integer)
        )
weather
# pivot the element into tmax and tmin
weather <- weather %>% 
    pivot_wider(
        names_from = element, 
        values_from = value
        )
weather
# print summary
weather %>% summary()
```

##### Question 7. 

Load the built in "anscombe" data frame and use "pivot_longer()" to separate all the x and y columns and categorize them into 4 sets

```{r question-7-response}
anscombe %>% 
    pivot_longer(
        # get the columns to separate
        cols = (starts_with("x") | starts_with("y")), 
        # .value will name the cols x or y respective to their values
        names_to = c(".value", "set"), 
        # names_pattern will match the column pattern x or y, set 1 to 4
        names_pattern = "(x|y)([1-4])"
        )
```
