---
title: "Data Visualization"
author: "Tyler Kephart"
date: "2024-08-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
## The instructions say to use the *mtcars* dataset but the graphs show the data and labels from the *mpg* dataset. Therefore, I will use *mpg* dataset to best match the graphs provided in the instructions.

### Question 1. Using the ~~mtcars~~ *mpg* data set and ggplot2, recreate the following graph.
```{r question1}
# name plot of displ to hwy to reuse in q2 and q3, hence name q123
q123 <- ggplot(data = mpg, mapping = aes(x = displ, y = hwy))
# adding plot points and smoothed mean line without confidence interval (se)
q123 + geom_point() + geom_smooth(se = FALSE)
```

### Question 2. Using the ~~mtcars~~ *mpg* data set and ggplot2, recreate the following graph.
```{r question2}
# same plot from q1 but with color mapped to drv
q123 + geom_point(mapping = aes(color = drv)) +
    geom_smooth(se = FALSE)
```


### Question 3. Using the ~~mtcars~~ *mpg* data set and ggplot2, recreate the following graph.
```{r question3}
# same plot as q2 but with linetypes for drv
q123 + geom_point(mapping = aes(color = drv)) +
    geom_smooth(mapping = aes(linetype = drv), se = FALSE)
```


### Question 4. Using the ~~mtcars~~ *mpg* data set and ggplot2, recreate the following graph.
```{r question4}
# plot cty to hwy mileage points and color on class and jitter them 
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
    geom_point(mapping = aes(color = class), position = "jitter")
```


### Question 5. Using the ~~mtcars~~ *mpg* data set and ggplot2, recreate the following graph.
```{r question5}
# plot hwy to cyl points as orange triangles and facet on class into 3 rows
ggplot(data = mpg, mapping = aes(x = hwy, y = cyl)) +
    geom_point(shape = 17, color = "orange") +
    facet_wrap(~class, nrow=3)
```

