---
title: "First Assignment"
author: "Tyler Kephart"
date: "08/03/2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Please complete all the tasks listed below. After completing the tasks download the .Rmd file and upload in the peer review item for grading.

# Additionally please write text between the code chunks explaining what each code chunk is about.

1.a bulleted list with sub-bullets under at least one item (10 pt)

* main item 1
  + sub item 1
* main item 2

2.a numbered list (10 pt)

1. number 1
2. number 2

3.some words italicized (5 pt)

*Italy*

4.some words in bold print (5 pt)

**Brave**

5.two levels of headers (10 pt)

## Headband

May be called sweatband.

### Sweatpants

May be called sweats.

6.the results of some R code without showing the code (10 pt)

```{r no_show_code, echo=FALSE}
1 + 1 < 0
```

7.a table of text at least 3 columns wide and 3 rows long (10 pt)

| First Col | Second Col | Third Col |
|-----------|------------|-----------|
| Row1 Col1 | Row1 Col2  | Row1 Col3 |
| Row2 Col1 | Row2 Col2  | Row2 Col3 |
| Row3 Col1 | Row3 Col2  | Row3 Col3 |

8.a plot (show the code that generates the plot) (10 pt)

```{r example_plot}
x <- 1:10
y <- 1/x
plot(x,y)
```

9.your session info (as output from R) (10 pt)

```{r session_info}
sessionInfo()
```

10.Execution of .Rmd file (10 pt)

11.Upload and execution of knitted pdf (10 pt)

